Machine Learning Methods for Discourse Segmentation of E-Mail Based Clinical Communication

### 2018-03-09 paper-amia-2018/manuscript.tex
Intro updated by Alex

\title{Machine Learning Models for the Segmentation of Clinical Exchange}

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

\institutes{
$^1$Department of Computer Science, Wayne State University, Detroit, Michigan \\  
$^2$Director, Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, Tallahassee, Florida\\
$^3$Department of Public Health Sciences, Henry Ford Health System, Detroit, Michigan\\
$^4$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\noindent{\bf Abstract}
\textit{Communication science to understand clinical process like Motivational Interviews (MI) is limited by traditional qualitative coding methods. Qualitative coding of communication data is very resource-intensive and time-consuming process, requires auto coding. This study utilized e-coaching data where email is used to deliver motivation-enhancing coaching to encourage healthy eating. A critical step toward automatic annotation of communication coding process is the segmentation of text data. In this study, we transformed segmentation task into a classification task and developed several state-of-the-art machine learning models including Support Vector Machine, Naive Bayes, K-Nearest Neighbor (KNN), Recurrent Neural Networks by utilizing contextual, topic and punctuation mark features. Results indicate that KNN is the best model and achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'' classes, respectively. This study has a great implication to save money and accelerate the pace of identifying effective communication strategies.}

\section*{Introduction}
Communication science to understand clinical process like Motivational Interviews (MI) is limited by traditional qualitative coding methods. Motivational Interviewing (MI) is an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009ten,miller2009toward}. Patient ``change talk'', statements of intrinsic motivation about their desire, ability, reasons, need for and commitment to behavior change, is an established mediator of health behavior change\cite{apodaca2009mechanisms}. Qualitative coding of communication data has been traditionally performed manually with pre-defined codes by trained coders, which is a tedious resource-intensive task, requiring an iterative process of human (subjective) interpretation of the text.

Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. Machine learning-based classification methods have been successfully applied to a variety of analytical tasks on textual data including classification\cite{nigam2000text}, sentiment analysis\cite{wang2012baselines}, and digital forensics\cite{de2002language}. A few studies have been done for the automatic annotation of clinical interactions in the setting of MI intervention. Lacson et al.\cite{lacson2005automatic} applied AdaBoost\cite{freund1999short} classifier for annotating the interactions in hemodialysis phone dialog as Clinical, Technical,
Backchannel, and Miscellaneous categories. Our research group has recently applied several machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}.

Experimental data utilized in the above MI studies were prepared by transcribing audio conversation which involves a counselor, a patient, and some cases a caregiver\cite{hasan2016study,kotov2015interpretable}. However, this study utilized e-coaching data which have different structure and context compared to traditional clinical data. E-coaches use email to deliver motivation-enhancing coaching to encourage healthy eating, grounded in the principles of motivational interviewing. E-coaching data is comprised of email responses which are unsegmented, unlike more traditional dyadic clinical communication where segmentation occurs naturally due to its conversation nature. Discourse analysis on e-coaching text can help to understand the communication process in e-coaching text by revealing the socio-psychological characteristics of a patient\cite{siegfried1995therapeutic,kalchbrenner2013recurrent,pierre2016neural}. By utilizing e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process.

A significant barrier to fully automated the communication coding process is the segmentation of the text data. During traditional qualitative coding, coders determine where to segment text, that is, where one code ends and another begins. Automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange as a non-traditional format. Second, a text segment does not necessarily belong to the entire sentence or collection of sentences. One sentence can be segmented into multiple MI behaviors or several sentences may represent a single MI code. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain 1 and 3 sentences, respectively.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/segment-example.png}
    \caption{\textbf{Segmentation of e-coaching text depicts the main challenges of boundary detection}.}
    \label{fig:text-segment}
\end{figure}

In this paper, we address the text segmentation problem by developing several state-of-the-art machine learning models to promote the automatic identification of best communication strategies without human interference. More specifically, we developed Support Vector Machine (SVM), Naive Bayes (NB), K-Nearest Neighbor (KNN), Long Short Term Memory (LSTM), and Gated Recurrent Unit (GRU) by utilizing contextual, topic and punctuation mark features, to find the best model for the segmentation of e-coaching text.

Previous studies mainly focus on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2016sentence} in the medical domain. Apostolova et al.\cite{apostolova2009automatic} applied SVM by utilizing word-vector cosine similarity metric combined with several heuristics to classify clinical report into semantic sections such as demographics, history, exam procedure, finding, impression, etc. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for the section classification. In 2009, Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing technique, terminology-based rule, and naive Bayesian score for identifying sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM exploiting with linear kernel and recurrent convolutional neural networks with prosodic, part of speech features and word embeddings, were trained by Kreuzthaler et al.\cite{kreuzthaler2015detection} and Griffis et al.\cite{griffis2016quantitative}, respectively, for the detection of sentence boundary.

Recently, an online clinical intervention called MENU GenY\cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) was proposed to test its efficacy. MENU GenY is a technology-based public health intervention to encourage increased fruit and vegetable intake among young adult age 21-30, utilized personalized e-coaching data. The goal of MENU GenY was to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical conversation in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still ignored while relying on manual hand-coded approach. Therefore, this study introduces a novel approach and the authors are not aware of any other work this approach has been considered for the segmentation of e-coaching text.

@@

\bibliographystyle{vancouver}
\bibliography{references}

>>>>

\title{Machine Learning Methods for Discourse Segmentation of E-Mail Based Clinical Communication}

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

\institutes{
$^1$Department of Computer Science, Wayne State University, Detroit, Michigan \\  
$^2$Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, Tallahassee, Florida\\
$^3$Department of Public Health Sciences, Henry Ford Health System, Detroit, Michigan\\
$^4$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\noindent{\bf Abstract}
\textit{Communication science to understand clinical process like Motivational Interviews (MI) is limited by traditional qualitative coding methods. Qualitative coding of communication data is very resource-intensive and time-consuming process, which requires auto coding. This study utilized e-coaching data where email is used to deliver motivation-enhancing coaching to encourage healthy eating. A critical step toward automatic annotation of communication coding process is the segmentation of text data. In this study, we transformed segmentation task into a classification task and developed several state-of-the-art machine learning models including Support Vector Machine, Naive Bayes, K-Nearest Neighbor (KNN), Recurrent Neural Networks by utilizing contextual, topic and punctuation mark features. Results indicate that KNN is the best model and achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'' classes, respectively. This study has a great implication to save money and accelerate the pace of identifying effective communication strategies.}

\section*{Introduction}
The emergence of e-Health technologies opened up new ways to deliver a variety of behavioral interventions to any demographic group of patients in any geographical location. Motivational interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009ten,miller2009toward}, is one type of these interventions. MI sessions are generally aimed at eliciting ``change talk'', or statements of intrinsic motivation about patients' own desire, ability, reasons, need for and commitment to behavior change, which have been established by previous research\cite{apodaca2009mechanisms} as a reliable mediator of health behavior change. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods.

Qualitative coding of motivational interviews with pre-defined codes has been traditionally performed manually by trained annotators, which is a tedious and resource-intensive process that involves several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis\cite{wang2012baselines}. In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and analysis \cite{hasan2018predicting} of in-person MI sessions. Specifically, we demonstrated that machine learning methods can be utilized for annotation of MI transcripts according to a simple communication code scheme with the accuracy comparable to human coders\cite{hasan2016study}. Experimental data utilized in these studies, however, were prepared by transcribing audio conversations, which were clearly segmented into utterances by a counselor, a patient, and, in some cases, a caregiver.

In this study, we focus on the analysis of e-Coaching sessions, MI interventions that are delivered via email and grounded in the principles of motivational interviewing. Specifically, the e-Coaches involved in this study used emails to communicate motivation-enhancing messages that encourage healthy eating among GenY adolescents. e-Coaching data is comprised of email responses, which are free-text documents, unlike more traditional dyadic clinical interviews that are naturally segmented into utterances due to their conversational nature.

The unstructured nature of e-Coaching exchanges poses a unique set of challenges for their qualitative analysis. A significant barrier to fully automating the behavior coding process of e-Coaching emails is their segmentation into textual fragments that correspond to distinct communication behaviors. Automating this task is a unique and challenging problem due to the following major reasons:

\begin{enumerate}
\item Emails are unstructured text that contains informal information exchange in a non-traditional format.
\item Discourse segments in e-Coaching emails do not necessarily correspond to sentences or collection of sentences. One sentence can be segmented into multiple MI behavior fragments. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

Figure~\ref{fig:text-segment} illustrates a segmentation of an e-Coaching email exchange, in which the first sentence is segmented into 2 MI behavior fragments, while the fourth and fifth segments correspond to one and three sentences, respectively. Segmentation of e-Coaching emails corresponds to a special type of discourse analysis \cite{webber2012discourse} aimed at better understanding the effective e-Coaching communication strategies and revealing the unique socio-psychological characteristics of a patient.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/segment-example.png}
    \caption{\textbf{Example of e-Coaching emails segmented into fragments that correspond to MI behaviors of an e-Coach and a patient}.}
    \label{fig:text-segment}
\end{figure}

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced contextual, topic and punctuation features and experimented with both traditional supervised machine learning methods, such as Support Vector Machine (SVM), Naive Bayes (NB) and $k$-Nearest Neighbor (KNN) classifiers, and deep learning methods, such as Long Short Term Memory (LSTM) and Gated Recurrent Unit (GRU), to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2016sentence}. Apostolova et al.\cite{apostolova2009automatic} applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for section classification. Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules, and Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Segmentation of e-Coaching emails is different from traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

Recently, an online clinical intervention called MENU GenY \cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) was proposed and evaluated. MENU GenY is a technology-based public health intervention that relies on personalized e-coaching to encourage increased fruit and vegetable intake among young adults, aged 21-30. The goal of MENU GenY was to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical conversation in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still performed manually, which slows down qualitative analysis of these interventions. This study introduces a novel computational approach to address this problem and the authors are unaware of any other work that focused on the same problem.

@@

\bibliographystyle{vancouver}
\bibliography{manuscript}
